Experiences Gamifying Developer Adoption\\ of Practices and Tools

### 2014-03-11 paper_id51.tex
Updated conference info
updated ICSE conference info
Partway through comment addressing.  The first two comments are partially addressed and last 3 are.

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{International Conference on Software Engineering}{2014 Hyderbad, India}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

Sascha is a developer with years of experience maintaining code, who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE mainly out of habit.  When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.  Sascha tried the code navigation tools and practices that were part of the game and soon found himself at the top of the game's leaderboard.  After learning the new practices Sascha spends less time navigating code to accomplish his maintenance tasks.

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may not be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

@@

\section{Related Work}

The two areas of related work we discuss are research studies using other monitoring methods and studies applying gamification to other areas.

@@

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each other's rankings while the control group only sees their personal rankings.  Results, though inconclusive, show competition increases the desired results.  In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data was shared with and divided the questions between whether sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.  
The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

The team goal ranked second behind coworker recommendation.  This response indicated that developers would be motivated to use new tools and practices if that helped their team in a competition between development teams at ABB.  The result shows that team competition would be first or second on the list for 47\% of developers to motivate their use of practices or tools.

Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.  The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.   Badges and management mandate received the most negative written comments in the survey.  A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me. A badge will not really do anything for me.''  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice more effective developers use when maintaining programs.   Structured navigation fits the requirements of a practice contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.  Thus we selected structured navigation as the practice to demonstrate gamification.

@@

In another processing step, Blaze aggregates events along the time-line into sessions.  A session is an abstract concept grouping multiple events into a sequence with a beginning and an end.  Of the many possible ways to define sessions, we chose to define sequences as all the events that transpire before and during a set of edits to a specific file.  When the developer edits a new file, all the events after the last edit to the previous file are included in the new sequence for the new file.  For example, the developer edits file A, then navigates through files B and C finally landing at line 100 of file D where they make another edit.  All the events between the edit of file A and the edit of file D are considered part of the edit session ending with the edit of file D.  Multiple edits to file D that occur before editing a different file are included in the same sequence ending with the edits to file D.  Although not at the method level, this way of generating sequences is similar to how Robillard et al. generated sequences based on located methods for their study \cite{wbsnipes:Robillard2004How}.

\subsubsection{Observations About Drivers of Edit Session Duration}

We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio.

@@

Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.  The attribute search method we chose was Greedy Step-wise search, which selects attributes based on their correlation with the target, and stops adding attributes when the next attribute decreases the correlation.    The top 5 out of the 15 possible categories ranked by this process are as follows:
\begin{enumerate}[itemsep=0mm]
\item Unstructured navigation
\item Debug - events from running the debugger
\item Edit - events that modify the code
\item Other Actions - other events in visual studio that we do not sub-classify
\item Build - events from running a build
\end{enumerate}

@@

Following our data analysis, we conducted one-on-one feedback sessions with the pilot participants. Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management. The participants received the feedback in the form of one-on-one presentations of quantitative charts and written analysis of the information.  The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.

>>>>

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{ICSE}{'14, May 31 -- June 7, 2014, Hyderabad, India}
\CopyrightYear{14}
\crdata{978-1-4503-2768-8/14/05}
% --- End of Author Metadata ---

@@

Sascha is a developer with years of experience maintaining code, who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE mainly out of habit.  When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.  Sascha tried the code navigation tools and practices that were part of the game and soon found himself at the top of the game's leaderboard.  After participating in the game, Sascha used the new practices more often and felt recognized for his achievement.
%Deleted the following due to Review #1 comments who concluded the paper was about improving productivity using structured navigation. After learning the new practices Sascha spends less time navigating code to accomplish his maintenance tasks.

Toolsmiths and researchers are continuously finding ways to improve the work of software developers.  However, as new practices and tools become available, developers like Sascha may not be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

%Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may not be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

@@

\section{Related Work}
\label{sec:RelatedWork}
The two areas of related work we discuss are research studies using other monitoring methods and studies applying gamification to other areas.

@@

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each other's rankings while the control group only sees their personal rankings.  In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data was shared with and divided the questions between whether sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows, as the community with which the data is shared gets smaller, more developer are comfortable with sharing the data.  If we take scope as the main concept, and limit the sharing to ``My Team'', we can say that 90\% of respondents are comfortable with sharing either anonymous or non-anonymous data if the scope of data were limited to ``My Team''.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.  
The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

The second ranked motivator for using tools and practices, ``Team Goal'', shows the survey response of ``It elevated my team`s achievement level in a game among software engineering teams at ABB.''   This response indicated that 47\% of developers would be motivated to use new tools and practices if that helped their team in a competition between development teams.

Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.  The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.   Badges and management mandate received the most negative written comments in the survey.  A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me. A badge will not really do anything for me.''  We conclude from this result, supported by findings in \cite{MurphyHill2012Improving,wbsnipes:Hill2011Peer}, that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} (discussed in Related Work \ref{sec:RelatedWork}) identify structured navigation as a practice more effective developers use when maintaining programs.  Structured navigation fits the requirements of a practice contained in the IDE, has proven benefits, and is not specifically assigned task for a developer.  Thus we selected structured navigation as the practice to demonstrate gamification.

@@

In another processing step, Blaze aggregates events along the time-line into sessions.  A session is an abstract concept grouping multiple events into a sequence with a beginning and an end.  Of the many possible ways to define sessions, we chose to define sequences to contain all the events that transpire before and during a set of edits to a file.  When the developer edits a file different from the file last edited, all the events after the last edit to the previous file are included in the new sequence for the new file.  For example, the developer edits file A, then navigates through files B and C finally landing at line 100 of file D where they make another edit.  All the events between the edit of file A and the edit of file D are considered part of the edit session ending with the edit of file D.  When this edit is followed by multiple edits to file D, the additional edits are included in the same sequence ending with the edits to file D.  In other words, a sequence is defined to contain the set of actions that occur between  Although not at the method level, this way of generating sequences is similar to how Robillard et al. generated sequences based on located methods for their study \cite{wbsnipes:Robillard2004How}.

\subsubsection{Observations About Drivers of Edit Session Duration}
\label{sec:driversOfEditSession}
We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio.

@@

Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.  The attribute search method we chose was Greedy Step-wise search, which selects attributes based on their correlation with the target, and stops adding attributes when the next attribute decreases the correlation.    Defining the attributes as the sum of developer time spent by category in each edit session, the top 5 out of the 15 possible categories ranked by this process are as follows:
\begin{enumerate}[itemsep=0mm]
\item Unstructured navigation
\item Debug - events from running the debugger
\item Edit - events that modify the code
\item Other Actions - other events in visual studio that we do not sub-classify
\item Build - events from running a build
\end{enumerate}

@@

Following our data analysis, we conducted one-on-one feedback sessions with the pilot participants. Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management. The participants received the feedback in the form of one-on-one presentations of quantitative charts and written analysis of the information.  The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.   The categories were identified based on what would be most relevant to the developer where some categories match findings in the Weka analysis in section \ref{sec:driversOfEditSession}.   The ``Actions'' category identified by Weka is omitted in the chart because it is a collection of events not categorized at a level useful for developers.  The category for structured navigation shows the practice we are trying to encourage as the opposite of the more common unstructured navigation practice.
